---
title: "Chapter 2 Examples"
output: html_notebook
---



```{r}
# remove all elements for a clean start
rm(list=ls(all=TRUE))
```

### Load Packages
```{r}
require(sas7bdat)
library(nlme)
library(lme4)
library(emmeans)
library(here)
```

### Load Data
```{r}
pathDir  <- here()
pathCh2  <- file.path(pathDir,"chapters/02/data/SAS_Chapter2.sas7bdat")
dsL2   <- read.sas7bdat(pathCh2, debug=TRUE) 
ds <- dsL2
```

### column names
```{r}
names(ds)
```

### counts/structure of data
```{r}
str(ds)
```

### change sex and demgroup to factors
```{r}
ds$sexMW <- as.factor(ds$sexMW)
ds$demgroup <- as.factor(ds$demgroup)
```

### description (mean, median, etc.)
```{r}
summary(ds)
```

### show the breakdown of the data types
```{r}
table(ds$sexMW, ds$demgroup)
```

### culminative sum
```{r}
cumsum(table(ds$demgroup))
```

### EmptyMeansModel
```{r}
model_null <- nlme::gls(cognition ~ 1, data=ds, method="REML")
summary(model_null)
```

### Null Model Covariance Parameter Estimate
```{r}
summary(model_null)$sigma^2
```

### center variables
```{r}
ds$age85 <- ds$age - 85
ds$grip9 <- ds$grip - 9
```

### model age
```{r}
model_age85 <- nlme::gls(cognition ~ 1 + age85, data=ds, method="REML")
summary(model_age85)
ds$model_age85 <- predict(model_age85)
```

### add age Model Covariance Parameter Estimate
```{r}
summary(model_age85)$sigma^2
```

### compare null model versus adding age
```{r}
anova(model_null, model_age85)
```

### model of unadjusted age
```{r}
model_age <- nlme::gls(cognition ~ 1 + age, data=ds, method="REML")
summary(model_age)
summary(model_age)$sigma^2
```

### Model add grip
```{r}
model_add_grip <- nlme::gls(cognition ~ 1 + age85 + grip9, data=ds, method="REML")
summary(model_add_grip)
summary(model_add_grip)$sigma^2
```

### Model add sex (M=0, W=1)
```{r}
ds$sexMW <- factor(ds$sexMW, levels=c(0, 1))
model_add_sex_m0 <- nlme::gls(cognition ~ 1 + age85 + grip9 + sexMW, data=ds, method="REML")
summary(model_add_sex_m0)
summary(model_add_sex_m0)$sigma^2
```

### Model add sex (W=1, M=0)
```{r}
ds$sexMW <- factor(ds$sexMW, levels=c(1, 0))
model_add_sex_w0 <- nlme::gls(cognition ~ 1 + age85 + grip9 + sexMW, data=ds, method="REML")
summary(model_add_sex_w0)
summary(model_add_sex_w0)$sigma^2
```

### Model add demgroup
```{r}
model_add_demgroup <- nlme::gls(cognition ~ 1 + age85 + grip9 + sexMW + demgroup, data=ds, method="REML")
summary(model_add_demgroup)
summary(model_add_demgroup)$sigma^2
```

### get all contrasts
```{r}
emmeans(model_add_demgroup, pairwise~demgroup)
```

### Model add interaction age85*grip9
```{r}
model_add_int <- nlme::gls(cognition ~ 1 + age85*grip9 + sexMW + demgroup, data=ds, method="REML")
summary(model_add_int)
summary(model_add_int)$sigma^2
```

### Model add interaction sex*dem
```{r}
model_add_ints <- nlme::gls(cognition ~ 1 + age85*grip9 + sexMW*demgroup, data=ds, method="REML")
summary(model_add_ints)
summary(model_add_ints)$sigma^2
```

